\documentclass[11pt]{article} 

\usepackage{amssymb,amsmath}

\newcommand{\numpy}{{\tt numpy}}            % tt font for numpy
\newcommand{\scipy}{{\tt scipy}}            % tt font for scipy
\newcommand{\matplotlib}{{\tt matplotlib}}  % tt font for matplotlib

\topmargin -1in
\textheight 9in
% \oddsidemargin  -.25in
% \evensidemargin -.20in
% \textwidth 7in

% \topmargin -.5in
% \textheight 7.5in
\oddsidemargin  -0.25in
\evensidemargin -0.20in
\textwidth 7in

\begin{document}

$$\mbox{\Large \bf CS 111: Homework 1: Due by 11:59 pm Sunday, January 15, 2023}$$
\par\smallskip\noindent
{\bf Submit your paper as one PDF file,
and tell GradeScope which page(s) each problem is on.}

\par\medskip\noindent
{\bf You will get full credit for this homework just by handing it in.}
The purpose of this homework is to review some of the math background
for the class, and to help you decide if there is anything you want to 
brush up on.  The t.a.s will answer questions about the quiz in discussion 
section on January 18.


\begin{enumerate}

\item
Let
$$ A = \left(
\begin{array}{ccc}
3 & -1 & 2 \\ 	
0 & 1 & 2 \\ 
1 & 0 & -1 \\
\end{array} 
\right). $$
What is $A^T$?  What is $A^2$?  What is $A^TA$?
(Do these computations both by hand and with \numpy.)

\item
The notation $||x||_2$ means the Euclidean norm (also called the 2-norm,
or just the length) of the vector $x$.  
What is $||(3, 1, 4, 1, 5)^T||_2$?

\item
Consider the following system of three equations in three unknowns.
\begin{eqnarray}
 2x_0 - 3x_1 + x_2  & =  & 1 \\
 2x_1 + 3x_2  & =  & 7 \\
 x_0 + x_2  & = & 4 
\end{eqnarray}
First write this system in the form $Ax=b$, 
where $A$ is a matrix and $x$ and $b$ are vectors. 
Second, write two lines of \numpy\ code that use {\tt np.array()} to
create $A$ and $b$ as \numpy\ arrays.

\item
What vector $x$ solves the system above? 

\item
Write down a 2-by-2 matrix $A$ and a 2-vector $b$ such that $Ax=b$ has no 
solution.  Explain in a sentence why there are no solutions.

\item
Write down a 2-by-2 matrix $A$ and a 2-vector $b$ such that $Ax=b$ has 
more than one solution.  Give two different solutions.

\item
Is there a 2-by-2 matrix $A$ and a 2-vector $b$ such that $Ax=b$ has
exactly two solutions?  Why or why not?

\item
Recall that a number $\lambda$ is an {\em eigenvalue} 
of a matrix $A$ if there is some vector $x$ 
(an {\em eigenvector}) for which $Ax = \lambda x$.
Give an eigenvalue of the following matrix, and a corresponding eigenvector.
$$ A = \left(
\begin{array}{cc}
4 & -1 \\ 	
-1 & 4 \\
\end{array} 
\right). $$

\item
If $f(x)=7x^3-2x^2+4x-5$, what is $f'(x)$, the derivative of $f(x)$?

\item
If $z = xe^{y/2}$, what is $\partial z / \partial x$?
What is $\partial z / \partial y$?

\item
If $f'(x) = x^2 + \sin x$, what is $f(x)$?

\item
The height in feet of a bullet fired straight up is
given by $h=1280t-16t^2$, where $t$ is in seconds.
What is the maximum height the bullet will reach? 
When will it hit the ground?

\item
Suppose $y$ is a function of $x$ that satisfies
$dy/dx = xy$, and also suppose $y=1$ when $x=0$.
Write $y$ as a function of $x$.

\end{enumerate}

\end{document}
